---
title: "Background Information"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
always_allow_html: true
output:
  html_fragment:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results='asis', echo=FALSE, warning=FALSE, message = FALSE)
# needed for trouble shooting
boo_DEBUG <- FALSE
if(boo_DEBUG==TRUE){
  # myConfig <- file.path(system.file(package="ContDataQC"), "extdata", "config.ORIG.R")
  # source(myConfig)
}## IF ~ boo_DEBUG ~ END
```

# Background
The Illinois Environmental Protection Agency (IEPA) is responsible for sampling and assessing Illinois’ surface water quality pursuant to the Clean Water Act (CWA) and identifying waterbodies that are not meeting water quality criteria. In addition to assessments based on water chemistry and physical parameters, IEPA developed a multimetric Index of Biotic Integrity (IBI; Karr et al. 1984) for lake assessments in 2015. Biological assessment, numerically represented by an IBI, is an evaluation of the biological condition of a water body using information on the structure and function of resident biological assemblages (U.S. EPA 1998). A biological index provides a measure for determining biological integrity that can be used in conjunction with other chemical or physical measurements to assess, manage, and protect aquatic life uses in lakes.

IEPA developed a lake macroinvertebrate IBI in 2015 that is applicable to lakes throughout the state of Illinois (Jessup et al. 2015). The process for index development included the designation of lake disturbance, establishment of lake classes based on natural variability, calibration of multiple macroinvertebrate metrics, and the combination of the most responsive metrics into an index. Each metric was selected and included in the index if it showed a consistent response along a known disturbance gradient. The responsiveness of metrics was evaluated within three site classes that separated the state of Illinois from north to south.

Since the IBI was calibrated, IEPA has continued to sample lakes, and in 2021 wanted to validate the IBI by applying it to newly collected data. However, in 2016, IEPA began testing a new sampling technique in which multiple lake habitats are sampled and combined in the field before sample processing in a laboratory. The new method expedites sample collection and processing time and is less costly than the previous method. This new method, termed Composite F, is slightly different from the sampling method that was used to collect samples in the 2015 IBI calibration dataset, termed Littoral ABC. The Littoral ABC sampling method is different from Composite F in that the multiple lake habitat samples are independently processed in the laboratory prior to a virtual composition. These methodological differences result in samples that are not directly comparable due to differences in the number of organisms that constitute a sample.

Although the 2015 index was validated using the Littoral ABC samples, the index performs less well on Composite F samples. A decrease in performance was unsurprising and can likely be attributed to the fact that fewer organisms are encountered in Composite F samples than Littoral ABC samples. Based on these results, the IEPA working group decided that the best course of action would be to forgo index validation in favor of a full index recalibration. IEPA identified that the methodological shift to Composite F sampling was more important for the biomonitoring program in general, and therefore, recognized a need to develop a new IBI purposefully built for such methods and sample sizes. This R Shiny app calculates the new lake macroinvertebrate IBI for use in Illinois lakes.

Reports detailing the development of the macroinvertebrate IBI can be found [here](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IEPAtools/Reports).

## Intended application
The IBI is intended to be applied to samples collected from freshwater lakes in Illinois using IEPA's Composite F sampling method. Composite F is a composite method wherein IEPA samples the three littoral habitats of a lake and then physically composites the samples together in the field. Laboratory sample processing includes a random subsample to a target of 500 organisms. For metric calculation, subsampled enumerations were multiplied up to whole sample densities based on the subsampling fraction in each habitat type (see Jessup et al. 2015 for details). This does not affect taxa richness or proportional composition within a single habitat sample. If you apply the IBI to samples that do not meet the criteria described above, results should be interpreted with caution.

## Input metrics and scoring formulae
Macroinvertebrate IBIs were calibrated for three regions - the Northern, Central, and Southern site classes (shown in Figure 1). Table 1 shows the input metrics and scoring formulae for the three IBIs. 

### Table 1. Metrics and scoring formulae for the IEPA IBI.
```{r table_1}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab1.dir <- "Bug_Metrics"

table1 <- read_excel(file.path(table.dir, table.file), sheet = tab1.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table1 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```
